Monitoring forest health using hyperspectral imagery: Does feature selection improve the performance of machine-learning techniques?

### 2021-06-24 docs/00-manuscripts/ieee/ieee-manuscript.tex
update methods section

To use the full potential of the hyperspectral data, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI}s were calculated.

NRIs follow the \ac{OMNBR} concept of data-driven information extraction from narrow-band indices of hyperspectral data \cite{thenkabail2000,thenkabail2018}.
While various index formulations such as band ratios or normalized ratios are available, indices involving the same bands are strongly correlated.
Since the widely-used NDVI index belongs to the family of normalized ratio indices (NRIs), which are implemented in the \texttt{hsdar} R package, we used the following NDI formula to combine all pairs of reflectances:

@@

To account for geometric offsets within the hyperspectral data, which were reported with up to one meter from \ac{ICGC}, a buffer of one meter around the centroid of each tree was used when extracting the reflectance values.
A pixel was considered to fall into a tree's buffer zone if the centroid of the respective pixel was touched by the buffer.
All of those pixels formed the final reflectance value of a single tree and were used as the base information to derive all additional feature sets.
In total, \(\frac{121*122}{2} = 7471\) NRIs were calculated.

@@

To reduce the feature space of a dataset, conceptually differing approaches exist: wrapper methods, filters, penalization methods (lasso and ridge) or \ac{PCA} \cite{bommert2020, das2001, guyon2003, jolliffe2016}.
In contrast to wrapper methods, filters can be added to the hyperparameter optimization step and have a lower computational footprint.
Due to the focus on filter methods in this manuscript, only this sub-group of feature selection methods will be introduced in greater detail in the following subsections.

@@

% Filter methods
The concept of filters originates from the idea of ranking features following a score calculated by an algorithm \cite{guyon2003}.
Some filter methods can only deal with specific types of variables (numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase of the model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a additional hyperparameter of the model.
The goal is to optimize the number of ranked features to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' can enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in ML which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was applied \cite{drotar2017a}.
Its final feature order is determined by the sum of all single filters ranks.

@@

\begin{itemize}
	\item  Extreme Gradient Boosting (XGBoost)
	\item  Random Forest (RF)
	\item  Penalized Regression (both L1 (lasso) and L2 (ridge))
	\item  Support Vector Machine (SVM, RBF Kernel)
\end{itemize}

RF and SVM are well established algorithms widely used in (environmental) remote sensing.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) has shown promising results in benchmark studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge does not remove variables from the model (penalization to zero) but just shrinks them to effectively zero, keeping them in the model.

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.
The selected hyperparameter settings are shown in appendix \autoref{tab:hyperparameter_limits}.
In addition the code which generated the matrix is available in the research compendium of this study (\url{10.5281/zenodo.2635403}).

@@

Some individual features were finally removed from these feature sets if they were numerically equivalent to another feature based on their pairwise correlation being greater than $1 - 10^{-10}$.
This preprocessing step reduced the number of covariates to 122 (HR), 86 (VI) and 7467 (NRI).

@@

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is a substantial reduction of the tuning budget that is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.
Tuning ranges are shown in \autoref{tab:hyperparameter_limits}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage \cite{binder2020}.
For \ac{PCA}, the number of principal components was tuned instead.
%The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_{try} = p^x_{try}$.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_\textrm{try} = p_\textrm{sel}^t$ as a function of the number of selected features, $p_\textrm{sel}$.
It was thus tuned on a logarithmic scale by varying $t$ between 0 (i.e. $m_\textrm{try} = 1$) and 0.5 (i.e. $m_\textrm{try}=\sqrt{p_\textrm{sel}}$).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features left after optimizing the feature percentage during tuning.

@@

All tasks of this study were conducted using the open-source statistical programming language R \cite{rcoreteam2019}.
A complete list of all R packages used in this study can be found in linked repositories.
Due to space limitations only the selected packages with high impact on this work will be explicitly cited.

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
Besides the availability of code and manuscript sources, a static webpage is available at (\url{https://github.com/pat-s/2019-feature-selection}), listing more side-analyses that were carried out during the creation of this study.

@@

\subsection{PCA of feature sets}

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
Overall, the response variable \enquote{tree defoliation} could be modeled with an \ac{RMSE} of 28 percentage points (p.p.).
SVM showed no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to five p.p (\autoref{fig:perf-result}).
SVM showed the best overall performance with a mean difference of around three p.p to the next best model (XGBoost) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 p.p. for learner SVM (without filter) but up to 54.3 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).

>>>>

To use the full potential of the hyperspectral data, all possible vegetation indices supported by the R package hsdar (89 in total) as well as all possible \ac{NRI} combinations were calculated.
NRIs follow the \ac{OMNBR} concept of data-driven information extraction from narrow-band indices of hyperspectral data \cite{thenkabail2000,thenkabail2018}.
While various index formulations such as band ratios or normalized ratios are available, indices involving the same bands are strongly correlated.
Since the widely-used NDVI index belongs to the family of normalized ratio indices (NRIs), which are implemented in the \texttt{hsdar} R package, we used the following normalized difference index (NDI) formula to combine all pairs of reflectances:

@@

To account for geometric offsets within the hyperspectral data, which were reported to be potentially up to one meter from \ac{ICGC}, a buffer of one meter around the centroid of each tree was used when extracting the reflectance values.
A pixel was considered to fall into a tree's buffer zone if the centroid of the respective pixel was touched by the buffer.
The pixel values of such were averaged and formed the final reflectance value of a single tree and were used as the base information to derive all additional feature sets.
In total, \(\frac{121*122}{2} = 7471\) NRIs were calculated.

@@

To reduce the feature space of a dataset, conceptually differing approaches exist: wrapper methods, filters, penalization methods (lasso and ridge) or \ac{PCA} \cite{bommert2020, das2001, guyon2003, jolliffe2016}.
In contrast to wrapper methods, filters can be added to the hyperparameter optimization step and have a lower computational footprint.
Due to the focus on filter methods in this manuscript, only this sub-group of feature selection methods will be introduced in greater detail in the following sections.

@@

% Filter methods
The concept of filters originates from the idea of ranking features following a score calculated by an algorithm \cite{guyon2003}.
Some filter methods can only deal with specific types of variables (e.g. numeric or nominal).
Filters only rank features, they do not decide which covariates to drop or keep \cite{drotar2015}.
The selection which features to keep for model fitting is usually done within the optimization phase during model fitting, along with the hyperparameter tuning.
Essentially, the number of covariates in the model is treated as a additional hyperparameter of the model.
The idea is to optimize the number of ranked features to the point at which the model achieves the best performance.

% Ensemble filter methods
Besides the concept of choosing a specific filter method to rank variables, studies showed that combining several filters using statistical operations such as 'minimum' or 'mean' can possibly enhance the predictive performance of the resulting models, especially when applied to multiple datasets \cite{abeel2010, drotar2017a}.
This approach is referred to as 'ensemble filtering' \cite{dietterich2000}.
Ensemble filters align with the recent rise of the 'ensemble' approach in ML which uses the idea of stacking to combine the predictions of multiple models, aiming to enhance predictive performance \cite{polikar2012, feurer2015, bolon-canedo2019}.
In this work the 'Borda' ensemble filter was used \cite{drotar2017a}.
Its final feature order is determined by the sum of all single filters ranks.

@@

\begin{itemize}
	\item  Extreme Gradient Boosting (XGBoost)
	\item  Random Forest (RF)
	\item  Penalized Regression (L1 (lasso) and L2 (ridge))
	\item  Support Vector Machine (SVM, RBF Kernel)
\end{itemize}

RF and SVM are well established algorithms and widely used in (environmental) remote sensing.
Extreme Gradient Boosting (commonly abbreviated as XGBoost) has shown promising results in benchmark studies in recent years.
Penalized regression is a statistical modeling technique capable of dealing with highly-correlated covariates by penalizing the coefficients of the model \cite{hastie2001}.
Common penalties are 'lasso' (L1) and 'ridge' (L2).
Ridge does not remove variables from the model (penalization to zero) but just shrinks them to effectively zero, keeping them in the model.

In total the benchmarking grid consisted of 156 experiments (6 feature sets $\times$ 3 ML algorithms $\times$ 8 feature-selection methods and for the L1/L2 models, 6 feature sets $\times$ 2 models.
The selected hyperparameter settings are shown in appendix \autoref{tab:hyperparameter_limits}.
In addition code and data are available in the research compendium of this study (\url{https://doi.org/10.5281/zenodo.2635403}) to support open and reproducible science.

@@

Some individual features of such were removed before using the datasets for modelling when being numerically equivalent to another feature based on the pairwise correlation being greater than $1 - 10^{-10}$.
This preprocessing step reduced the number of covariates to 122 (HR), 86 (VI) and 7467 (NRI).

@@

In this work, an initial design of 30 randomly composed hyperparameter settings in combination with a termination criterion of 70 iterations was used, resulting in a total budget of 100 evaluated hyperparameter settings per fold.
The advantage of this tuning approach is a substantial reduction of the tuning budget that is required to find a setting close to the global optimization minimum.
\ac{MBO} may outperform methods that do not use information from previous iterations, such as random search or grid search \cite{bergstra2012}.
Tuning ranges used in this work are shown in \autoref{tab:hyperparameter_limits}.

To optimize the number of features used for model fitting, the percentage of features was added as a hyperparameter during the optimization stage \cite{binder2020}.
For \ac{PCA}, the number of principal components was tuned.
%The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_{try} = p^x_{try}$.
The RF hyperparameter \texttt{\(m_{try}\)} was re-expressed as $m_\textrm{try} = p_\textrm{sel}^t$, a function of the number of selected features, $p_\textrm{sel}$.
It was thus tuned on a logarithmic scale by varying $t$ between 0 (i.e. $m_\textrm{try} = 1$) and 0.5 (i.e. $m_\textrm{try}=\sqrt{p_\textrm{sel}}$).
This was necessary to ensure that \texttt{\(m_{try}\)} was not chosen higher than the available number of features that were left after optimizing the feature percentage during tuning.

@@

All tasks of this study were conducted using the open-source statistical programming language R \cite{rcoreteam2019}.
A complete list of all R packages used in this study can be found in the linked repositories mentioned in the next paragraph.
Due to space limitations only the selected packages with high impact on this work will be explicitly cited.

The algorithm implementations of the following packages have been used: xgboost \cite{chen2016} (\textit{Extreme Gradient Boosting}), kernlab \cite{kernlab} (Support Vector Machine) and glmnet \cite{glmnet} (penalized regression).
The filter implementations of the following packages have been used: praznik \cite{praznik}, FSelectorRcpp \cite{fselectorrcpp}.
Package mlr \cite{mlr} was used for all modeling related steps.
drake \cite{drake} was used for structuring the work and reproducibility.
This study is available as a research compendium on Zenodo (\url{10.5281/zenodo.2635403}).
Besides the availability of code and manuscript sources, a static webpage is available at (\url{https://pat-s.github.io/2019-feature-selection}), listing more side-analyses that were carried out during the creation of this study.

@@

\subsection{PCA-based variance analysis of feature sets}

@@

% - RMSE around 27
% - MBO tuned penalized methods are substantially worse
Overall, the response variable \enquote{tree defoliation} could be modeled with an \ac{RMSE} of 28 percentage points (p.p.) (\autoref{fig:perf-result}).
SVM showed almost no differences in RMSE across feature sets whereas other learners (RF, SVM, XGBoost, lasso and ridge) differed up to five p.p (\autoref{fig:perf-result}).
SVM showed the best overall performance with a mean difference of around three p.p to the next best model (XGBoost) (\autoref{tab:best-learner-perf}).
Performance differences between test folds were large: Predicting on Luiando resulted in an RMSE of 9.0 p.p. for learner SVM (without filter) but up to 54.3 p.p. when testing on Laukiz2 (\autoref{tab:svm-single-fold-perf}).
